Interim report 2\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-08-22 interimreport2.tex
major revisions

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
%\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{todonotes}

@@

\begin{document}
\SweaveOpts{concordance=TRUE}

\title{Interim report 2\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\today  %\today is replaced with the current date \date{....}
\maketitle

@@

A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups. Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks. In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection. These additional analyses were:
\begin{itemize}
\item Multivariate modelling to account for confounding within the univariate analyses.
\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
\item Analyses of changing risk factors through time during outbreaks.
\item Modelling transmission chains in the population to understand effective reproduction number.
\item Development of risk maps to understand measles importation.
\end{itemize}

<<echo=FALSE>>=
## regression analyses - measles
## get data
# set wd
 setwd("~/Massey 2014/DHayman_20140627")
# read data
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
                 data = data , FUN=sum)
library(reshape)
library(reshape2)
library(plyr)
library(pscl)

test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
head(test)
dim(test)
caseyr<-aggregate( DiseaseName ~ AgeInYears, 
                   data = test , FUN=sum)
caseyr
test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
test$AgeInYears<-as.factor(test$AgeInYears)
tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
test$AgeInYears<-tage
#summary(test$EthnicityPrioritised)
teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
                                     "Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));

test$EthnicityPrioritised<-teth
head(test)
test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
test <- within(test, NZDep13[NZDep13 == "0"]<-NA)

# want to set NZ Deprivation to the appropriate year

testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
                              ifelse (test$RptYear == "2008",test$NZDep06,
                              ifelse (test$RptYear == "2009",test$NZDep06,
                              ifelse (test$RptYear == "2010",test$NZDep06,
                              ifelse (test$RptYear == "2011",test$NZDep06,
                              ifelse (test$RptYear == "2012",test$NZDep06,
                              ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
head(testt)
tail(testt)
str(testt)
testt$NZDep<-as.factor(testt$NZDep-1)

testtable<-aggregate( cbind( DiseaseName + as.numeric(SurvWeek)) ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
                 data = testt , FUN=sum)
dim(testtable)
colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
head(testtable)

tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
                                 "9"="6-10","10"="6-10"));

testtable$NZDep<-tnzd
head(testtable)
tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
                 data = testtable , FUN=sum)

### denominator data
 setwd("~/Massey_2014/measles/data")
denom<-read.csv("NZDep2006Denominators.csv",header=T)
head(denom)
denom<-denom[,-c(9)]
summary(denom)
dim(denom)
denom<-denom[,1:18]
row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
sum(row.has.na)
denomt <- denom[!row.has.na,]
head(denomt)
denomt<-denomt[denomt$Sex_code==99,]
denomt$Eth_Level<-as.factor(denomt$Eth_Level)
denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
dm<-cbind(denomt[,c(3,6,9:18)])
head(dm)
dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
dm$Age_Label<-dage
colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
head(dm)
datadm<-dm[dm$Age=="<6",]
dataA<-(rbind(datadm,datadm)/2)
dataA<-round(dataA[,3:12])
dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
ddm<-dm[!dm$Age=="<6",]
dm<-rbind(dataA,ddm)
dm$Ethnicity <- factor(dm$Ethnicity)
deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
                              "Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
dm$Ethnicity<-deth
head(dm)

popn<-melt(dm,id.vars=c("Age","Ethnicity"),measure.vars=c("Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10"))
colnames(popn)<-c("Age","Ethnicity","NZDep","Popn")
popn
pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
                                 "Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
                                 "Dep9"="6-10","Dep10"="6-10"));

popn$NZDep<-pnzd
head(popn)
tp<-aggregate( cbind(Popn) ~ NZDep+Age+Ethnicity, 
               data = popn , FUN=sum)

tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
testtable$merge <- paste(testtable$NZDep, testtable$Age, testtable$Ethnicity)
testtable <- testtable[testtable$Ethnicity!="None",]

tp$cases <- 0
cases <- matrix(0, length(tp$merge),1)
rownames(cases) <- tp$merge
cases[testtable$merge,] <- testtable$Cases
tp$cases <- cases

## plot this figure?
hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
tp$Ethnicity<- relevel(tp$Ethnicity, "European")
model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tp,family="quasipoisson")
summary(model)

model1<-update(model,~.-Age:Ethnicity:NZDep)
summary(model1)
anova(model1,test="F") # F test, not Chisq, because dispersion estimated by moments

model2<-update(model1,~.-Age:NZDep)
summary(model2)
anova(model2,test="F")

par(mfrow=c(2,2))
hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
res<-predict(model2)
plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
cor(exp(res),tp$cases)
cor.test(exp(res),tp$cases)
hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
par(mfrow=c(1,1))
plot(tp$cases/tp$Popn,main="Cases per capita",ylab="Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
     xlab="Category")
legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
       col=c(1:4,6),pch=16,bty="n")

## drop MLA
tpsub<-tp[!(tp$Ethnicity=="MLA"),]

hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)

model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
summary(model)

model1<-update(model,~.-Age:Ethnicity:NZDep)
summary(model1)
anova(model1,test="F")

model2<-update(model1,~.-Ethnicity:NZDep)
summary(model2)
anova(model2,test="F")

model3<-update(model2,~.-Age:NZDep)
summary(model3)
anova(model3,test="F")

##

par(mfrow=c(2,2))
hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
plot(tpsub$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
res<-predict(model3)
plot(exp(res),tpsub$cases,xlab="predictions",ylab='cases',main="Fit",pch=16,col="darkgrey")
cor(exp(res),tpsub$cases)
cor.test(exp(res),tpsub$cases)
#abline(lm(exp(res)~tpsub$cases))
hist(model3$residuals,main="Histogram of residuals",xlab="residuals",col="grey")

##
@

@@

@@

The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$150 million \citep{carabin3]. The estimated cost for a case ranged from US\$189--344 \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755 \citep{carabin2}. \citep{stack11} estimated the economic benefits from cases averted due to measles vaccination. They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion of costs averted between 2011 and 2020. Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining \<1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.

@@

<<echo=FALSE>>=
costtable1<-read.csv("costtable1.csv",header=T)
@

\begin{table}
\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(costtable1, file="", table.env=FALSE,rowname=NULL)
@
\label{table:direct}
\end{table}

The total cost for the 192 publicly funded hospital discharges with a measles primary diagnosis was \$289,694 (Table~\ref{table:hosp}). The mean cost per case was \$1,509. The mean cost per day of stay in the hospital was \$1,154.

<<echo=FALSE>>=
costtable2<-read.csv("costtable2.csv",header=T)
@

\begin{table}
\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2008--2013}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(costtable2, file="", table.env=FALSE,rowname=NULL)
@
\label{table:hosp}
\end{table}

From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 -- 19 June 2014). The number of cases by age group is shown in Table~\ref{table:freq}. Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 47\% and 33\%, respectively.

<<echo=FALSE>>=
costtable3<-read.csv("costtable3.csv",header=T)
@

\begin{table}
\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(costtable3, file="", table.env=FALSE,rowname=NULL)
@
\label{table:freq}
\end{table}

The majority (191/192) of the measles cases stayed in the hospital from 0 to 7 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213. Eighty-three of the cases stayed for 0 days and the remaining cases stayed a mean of 2.30 (median 2) days.
The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2008--2013 appear in Table~\ref{table:cases}. The cost for these 192 hospitalisations was \$289,964. Of the 292 cases, 87 were female and 105 were male. The number of days spent in the hospital was 251, making the mean daily cost \$1,509. Of the 251 days, 114 were spent by females and 137 by males. The mean hospital stay per case was 1.31 and 1.30 days for females and males, respectively. The mean cost per case was \$1,573 and \$1,456 for females and males, respectively.

<<echo=FALSE>>=
costtable4<-read.csv("costtable4.csv",header=T)
@

\begin{table}
\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2008--2013}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(costtable4, file="", table.env=FALSE,rowname=NULL)
@
\label{table:cases}
\end{table}

Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}). Results showed an increase of approximately \$358 per day of hospitalisation, \$139 for a female case, and \$1 per year of case diagnosis. The cost was found to decrease by approximately \$5 per year of case age.

<<echo=FALSE>>=
costtable5<-read.csv("costtable5.csv",header=T)
@

\begin{table}
\caption{Regression results ($R^{2}_\textsf{adj} = 0.9$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=192$) in New Zealand, 2008 -- 2013}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(costtable5, file="", table.env=FALSE,rowname=NULL)
@
\label{table:regression}
\end{table}

Wages lost due to measles were calculated based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss. If the case were less than 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age. Total wage lost for the 192 cases and care givers was estimated to be \$166,638. This consisted of \$90,620 for the cases and \$76,018 for the care giver, but did not include wage losses for cases under 15 years of age.

This final figure brings an approximate estimate of \$786,479 for the current 2014 outbreak, which is comprised of earnings lost, case management and hospitalisation costs.

\subsection{Cost analysis discussion}
The results presented here are based on available data. While some of the data are complete and detailed, this is not true of all the data. In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed. For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals. In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals. In addition, similar data would be needed for non-publicly funded hospitals, e.g. private clinics. Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.

>>>>

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
%\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{todonotes}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{array}
\usepackage{rotating}
\usepackage{threeparttable}
\usepackage{blindtext}
\usepackage{dcolumn}
\usepackage{tabularx}

@@

\usepackage{Sweave}
\begin{document}
\input{interimreport2-concordance}

\title{Interim report 2\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\maketitle

@@

A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups. Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks. In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection. The additional analyses included in this section are:
\begin{itemize}
\item Multivariate modelling to account for confounding within the univariate analyses.
\item Analyses of changing risk factors through time during outbreaks.
\end{itemize}

\begin{Schunk}
\begin{Sinput}
> summary(model3)
\end{Sinput}
\begin{Soutput}
Call:
glm(formula = cases ~ Age + Ethnicity + NZDep + Age:Ethnicity + 
    offset(log(Popn)), family = "quasipoisson", data = tpsub)

Deviance Residuals: 
    Min       1Q   Median       3Q      Max  
-6.1174  -1.1562   0.0000   0.8599   6.7625

Coefficients:
                            Estimate Std. Error t value Pr(>|t|)    
(Intercept)               -8.268e+00  8.641e-01  -9.569 1.06e-08 ***
Age3-5                     6.931e-01  1.053e+00   0.658 0.518322    
Age6-17                    7.176e-01  9.064e-01   0.792 0.438300    
Age18-24                  -5.001e-01  1.149e+00  -0.435 0.668367    
Age25+                    -2.620e+00  1.158e+00  -2.262 0.035602 *  
EthnicityAsian             2.167e+00  1.203e+00   1.801 0.087576 .  
EthnicityMaori             1.772e+00  1.061e+00   1.671 0.111064    
EthnicityPacific           3.871e+00  9.405e-01   4.117 0.000587 ***
NZDep6-10                 -4.878e-01  2.608e-01  -1.870 0.076939 .  
Age3-5:EthnicityAsian     -2.431e-15  1.473e+00   0.000 1.000000    
Age6-17:EthnicityAsian    -2.473e+00  1.530e+00  -1.616 0.122649    
Age18-24:EthnicityAsian   -7.910e-01  1.622e+00  -0.488 0.631293    
Age25+:EthnicityAsian     -6.025e-01  1.792e+00  -0.336 0.740386    
Age3-5:EthnicityMaori     -2.845e+00  2.177e+00  -1.307 0.206941    
Age6-17:EthnicityMaori    -2.102e-01  1.121e+00  -0.187 0.853294    
Age18-24:EthnicityMaori   -8.881e-01  1.630e+00  -0.545 0.592102    
Age25+:EthnicityMaori     -1.090e+00  2.011e+00  -0.542 0.594088    
Age3-5:EthnicityPacific   -4.423e+00  2.581e+00  -1.714 0.102866    
Age6-17:EthnicityPacific  -2.253e+00  1.051e+00  -2.143 0.045264 *  
Age18-24:EthnicityPacific -8.321e-01  1.318e+00  -0.631 0.535272    
Age25+:EthnicityPacific   -2.718e-01  1.338e+00  -0.203 0.841210    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for quasipoisson family taken to be 16.26672)

    Null deviance: 3097.24  on 39  degrees of freedom
Residual deviance:  291.05  on 19  degrees of freedom
AIC: NA

Number of Fisher Scoring iterations: 6
\end{Soutput}
\end{Schunk}

\begin{Schunk}
\begin{Sinput}
> anova(model3,test="F")
\end{Sinput}
\begin{Soutput}
Analysis of Deviance Table

Model: quasipoisson, link: log

Response: cases

Terms added sequentially (first to last)

              Df Deviance Resid. Df Resid. Dev       F    Pr(>F)    
NULL                             39    3097.24                      
Age            4  1573.25        35    1523.98 24.1790 3.142e-07 ***
Ethnicity      3   738.20        32     785.78 15.1271 2.853e-05 ***
NZDep          1    55.39        31     730.39  3.4051   0.08064 .  
Age:Ethnicity 12   439.34        19     291.05  2.2507   0.05519 .  
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}
\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{Rplot_quasipoisson_noMLA.pdf}
     \caption{test}
     \label{fig:regn}
\end{figure}

@@

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{cases_by_dhb_2009_2014.pdf}
     \caption{Measles cases by district health board (DHB) from 2009 to 2014}
     \label{fig:dhbcases}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
     \label{fig:outbreaks}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{averageR0.pdf}
     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
     \label{fig:r0}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
     \caption{test}
     \label{fig:percap}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
     \caption{test}
     \label{fig:percap}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
     \caption{test}
     \label{fig:percap}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{AgeCases.pdf}
     \caption{test}
     \label{fig:agecases}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{AgeOfVaccinationVaccinated.pdf}
     \caption{test}
     \label{fig:ageofvacc}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{NumbersToVaccinate.pdf}
     \caption{test}
     \label{fig:numvacc}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{PopulationNaive.pdf}
     \caption{test}
     \label{fig:naive}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{simulations.pdf}
     \caption{test}
     \label{fig:sim}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{vaccinationstatusage.pdf}
     \caption{test}
     \label{fig:vaccstat}
\end{figure}

@@

The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$150 million \citep{carabin3}. The estimated cost for a case ranged from US\$189--344 \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755 \citep{carabin2}. \citep{stack11} estimated the economic benefits from cases averted due to measles vaccination. They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion of costs averted between 2011 and 2020. Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining <1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.

@@

\begin{table}
\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
%latex.default(costtable1, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllll}
\hline\hline
\multicolumn{1}{c}{Category}&\multicolumn{1}{c}{January}&\multicolumn{1}{c}{February}&\multicolumn{1}{c}{March}&\multicolumn{1}{c}{Total}\tabularnewline
\hline
PHN&55,296&71,175&24,087&150,558\tabularnewline
PHA&0&0&2,656&2,656\tabularnewline
Data support&0&7,752&4,552&12,304\tabularnewline
Supervisors&10,656&10,464&3,232&24,352\tabularnewline
IMT&32,918&28,624&7,156&68,698\tabularnewline
SSO&0&2,746&1,186&3,932\tabularnewline
Measles operations&1,800&10,326&6,678&18,804\tabularnewline
Operations partner&2,187&14,580&7,290&24,057\tabularnewline
IMT controller partner&2,916&14,580&7,290&24,786\tabularnewline
Total&105,773&160,247&64,127&330,147\tabularnewline
\hline
\end{tabular}\end{center}\label{table:direct}
\end{table}

The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hosptial was \$550,024 (Table~\ref{table:hosp}). The mean cost per case was \$1,877. The mean cost per day of stay in the hospital was \$1,170.

\begin{table}
\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
%latex.default(costtable2, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lrrlll}
\hline\hline
\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Days}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Per.case}&\multicolumn{1}{c}{Per.day}\tabularnewline
\hline
2000&$  6$&$ 13$&8,850&1,475&681\tabularnewline
2001&$ 13$&$ 18$&11,267&867&626\tabularnewline
2002&$  5$&$  2$&3,869&774&1,934\tabularnewline
2003&$  9$&$ 12$&10,241&1,138&853\tabularnewline
2004&$  4$&$  5$&4,765&1,191&953\tabularnewline
2005&$  3$&$ 11$&5,111&1,704&465\tabularnewline
2006&$  1$&$  0$&602&602&NC\tabularnewline
2007&$  5$&$ 25$&82,977&16,595&3,319\tabularnewline
2008&$  3$&$  1$&3,038&1,013&3,038\tabularnewline
2009&$ 29$&$ 38$&40,782&1,406&1,073\tabularnewline
2010&$  5$&$  5$&6,701&1,340&1,340\tabularnewline
2011&$132$&$189$&205,303&1,555&1,086\tabularnewline
2012&$ 19$&$ 12$&28,540&1,502&2,378\tabularnewline
2013&$  4$&$  6$&5,330&1,333&888\tabularnewline
2014&$ 55$&$133$&132,648&2,412&997\tabularnewline
TOTAL&$293$&$470$&550,024&1,877&1,170\tabularnewline
\hline
\end{tabular}\end{center}\label{table:hosp}
 \centering
 \begin{tablenotes}
      \small
      \item As of 11 July, 2014. NC - not calculated.
    \end{tablenotes}
\end{table}

From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before 1 January 2014, so 187 occurred from Jan 2013 -- 19 June 2014). The number of cases by age group is shown in Table~\ref{table:freq}. Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 47\% and 33\%, respectively.

\begin{table}
\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
%latex.default(costtable3, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lrrr}
\hline\hline
\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Admitted}&\multicolumn{1}{c}{Proportion}\tabularnewline
\hline
\textless  15 months&$ 21$&$10$&$0.47$\tabularnewline
15 months – 3 years&$  7$&$ 1$&$0.14$\tabularnewline
4 – 9 years&$  8$&$ 0$&$0.00$\tabularnewline
10 -1 19 years&$132$&$12$&$0.09$\tabularnewline
\textgreater  19 years&$ 33$&$11$&$0.33$\tabularnewline
Total&$201$&$34$&$0.17$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:freq}
\end{table}

The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 7 19 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213. Nearly 40\% (114/293) of the cases did not spend a night in the hospital, while approximately one-quarter (69/293) spent 1 night and more than three-quarters (222/293) spent less than three nights in the hospital. Only eight cases spent a week or more in the hospital. Due to the small number of cases spending a week or more in the hospital, the regression analysis to determine the association between cost of hospitalisation was limited to the 285 cases hospitalised for seven or fewer days. The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\ref{table:cases}.
